---
title: "advent_of_code_captcha"
author: "Paul Dong"
date: "2019/9/10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

The captcha requires you to review a sequence of digits (your puzzle input) and
find the sum of all digits that match the next digit in the list. The list is
circular, so the digit after the last digit is the first digit in the list.

For example:
    
- 1122 produces a sum of 3 (1 + 2) because the first digit (1) matches the 
second digit and the third digit (2) matches the fourth digit.
- 1111 produces 4 because each digit (all 1) matches the next.
- 1234 produces 0 because no digit matches the next.
- 91212129 produces 9 because the only digit that matches the next one is 
the last digit, 9.

```{r sol1}
# parse input
input_tidy <- function(input){
  str_split(input,"") %>% 
    unlist %>% 
    parse_number
}

compare <- function(input){
  tmp <- input_tidy(input) %>% 
    tibble(original = .) 
  tmp <- tmp %>% 
    mutate(lagged = lead(original, default = .$original[[1]])) 
  tmp <- tmp %>% 
    mutate(compare = original == lagged) %>% 
    summarise(sum = sum(compare, na.rm = T)) %>% 
    pull(sum) %>% 
    .[[1]]
  tmp
}
```

```{r test_sol1}
input_tidy("1122")
input_tidy("1111")

compare("1122")
compare("1111")
compare("1234")
compare("91212129")

```

Now, instead of considering the next digit, it wants you to consider the digit
halfway around the circular list. That is, if your list contains 10 items, only
include a digit in your sum if the digit 10/2 = 5 steps forward matches it.
Fortunately, your list has an even number of elements.

For example:

- 1212 produces 6: the list contains 4 items, and all four digits match the
digit 2 items ahead.
- 1221 produces 0, because every comparison is between a 1 and a 2.
- 123425 produces 4, because both 2s match each other, but no other digit has 
a match.
- 123123 produces 12.
- 12131415 produces 4.

```{r}
input_tidy <- function(input){
  str_split(input,"") %>% 
    unlist %>% 
    parse_number
}

compare <- function(input){
  tmp <- input_tidy(input) %>% 
    tibble(original = .) 
  steps <- nrow(tmp) / 2
  tmp <- tmp %>% 
    mutate(lagged = lead(original, steps))
  tmp$lagged[(steps+1):nrow(tmp)] <- tmp$original[1:steps]
  tmp <- tmp %>% 
    mutate(compare = original == lagged) %>% 
    summarise(sum = sum(compare)) %>% 
    pull(sum) %>% 
    .[[1]]
  tmp
}
```
```{r}
compare("1212")
compare("1221")
compare("123425")
compare("123123")
compare("12131415")
```
The spreadsheet consists of rows of apparently-random numbers. To make sure the
recovery process is on the right track, they need you to calculate the
spreadsheet's checksum. For each row, determine the difference between the
largest value and the smallest value; the checksum is the sum of all of these
differences.

For example, given the following spreadsheet:

5 1 9 5
7 5 3
2 4 6 8

The first row's largest and smallest values are 9 and 1, and their difference 
is 8.
The second row's largest and smallest values are 7 and 3, and their difference
is 4.
The third row's difference is 6.
In this example, the spreadsheet's checksum would be 8 + 4 + 6 = 18.

```{r}
data <- list(c(5, 1, 9, 5),
              c(7, 5, 3),
              c(2, 4, 6, 8))
```

```{r sol2}

max_min <- function(d){
  max(d) - min(d)
}

checksum <- function(data){
  map_dbl(data, max_min) %>% sum
}
```

```{r test2}
checksum(data)
```

It sounds like the goal is to find the only two numbers in each row where one evenly
divides the other - that is, where the result of the division operation is a whole
number. They would like you to find those numbers on each line, divide them, and add up
each line's result.

For example, given the following spreadsheet:

5 9 2 8
9 4 7 3
3 8 6 5
In the first row, the only two numbers that evenly divide are 8 and 2; the result of
this division is 4.
In the second row, the two numbers are 9 and 3; the result is 3.
In the third row, the result is 2.
In this example, the sum of the results would be 4 + 3 + 2 = 9.

What is the sum of each row's result in your puzzle input?
```{r sol2a}

data <- list("5 9 2 8", "9 4 7 3", "3 8 6 5")

data <- data %>%  
  map(~ str_split(.x," ") %>% unlist) %>% 
  map(parse_number)

even_divide <- function(d){
  d <- sort(d)
  l <- length(d)
  for(i in 1:(l-1)){
    divider <- d[i]
    dividee <- d[(i+1):l]
    r <- dividee / divider
    for (r_i in r){
      if (r_i == round(r_i)){
        return(r_i)
      }
    }
  }
}

checksum <- function(data){
  map_dbl(data, even_divide) %>% sum
}
```

```{r test2a}
checksum(data)
```
```{r}
data %>% as.matrix
which.max(data %>% array)
```


You come across an experimental new kind of memory stored on an infinite two-dimensional
grid.

Each square on the grid is allocated in a spiral pattern starting at a location marked 
1 and then counting up while spiraling outward. For example, the first few squares are
allocated like this:

17  16  15  14  13
18   5   4   3  12
19   6   1   2  11
20   7   8   9  10
21  22  23---> ...
While this is very space-efficient (no squares are skipped), requested data must be
carried back to square 1 (the location of the only access port for this memory system)
by programs that can only move up, down, left, or right. They always take the shortest
path: the Manhattan Distance between the location of the data and square 1.

For example:

Data from square 1 is carried 0 steps, since it's at the access port.
Data from square 12 is carried 3 steps, such as: down, left, left.
Data from square 23 is carried only 2 steps: up twice.
Data from square 1024 must be carried 31 steps.
How many steps are required to carry the data from the square identified in your puzzle
input all the way to the access port?

```{r}
Register <- R6::R6Class("Register", list(
  value = NULL,
  goal = NULL,
  fin_i = NULL,
  fin_j = NULL,
  i = NULL,
  j = NULL,
  initialize = function(value, goal, i, j){
    self$value <-  value
    self$goal <- goal
    self$i <- i
    self$j <- j
    invisible(self)
  },
  increase = function() {
    if(self$value == self$goal) {
      self$fin_i = self$i
      self$fin_j = self$j
    }
    self$value <- self$value + 1 
    invisible(self)
  }
))
```

```{r}
calc_steps <- function(n){
  # an issue with the end of the series
  n_sqrt <- n %>% sqrt
  if (n_sqrt == round(n_sqrt) && (n_sqrt%%2!=0)) return (n_sqrt - 1)
  
  r_max_1 <- n %>% sqrt %>% floor #r_max - 1
  if(r_max_1%%2==0) r_max_1 <- r_max_1 - 1
  r_max <- r_max_1 + 2 # size of the matrix
  
  M <- matrix(NA, nrow = r_max, ncol = r_max) # for ease of debugging
  
  value <- r_max_1 ^ 2 + 1 # start from the outermost layer
  init_i <- r_max - 1
  init_j <- r_max
  
  reg <- Register$new(value, n, init_i, init_j)
  
  for(iter in reg$i:1){
    reg$i <- iter
    M[reg$i,reg$j] <- reg$value
    reg$increase()
  }
  
  for(iter in (reg$j-1):1){
    reg$j <- iter
    M[reg$i,reg$j] <- reg$value
    reg$increase()
  }
  
  for(iter in 2:r_max){
    reg$i <- iter
    M[reg$i,reg$j] <- reg$value
    reg$increase()
  }
  
  for(iter in 2:r_max){
    reg$j <- iter
    M[reg$i,reg$j] <- reg$value
    reg$increase()
  }
  
  c_i <- ceiling(r_max / 2)
  c_j <- ceiling(r_max / 2)
  
  result <- abs(reg$fin_i - c_i) + abs(reg$fin_j - c_j)
  result
}
```
Location of the center
```{r}
calc_steps(1)
```


